Inferring disease architecture and predictive ability\\with LDpred2-auto

### 2023-10-03 paper/paper-infer.tex
add reponses to the text

Most traits and diseases in humans are heritable.
What differs is the genetic architecture of each trait that can be parameterized by three key terms: the heritability (i.e.\ the proportion of phenotypic variation explained by genetics), the polygenicity (i.e.\ the fraction of genomic variants that have a non-zero effect on the trait), and the causal effect distribution (i.e.\ how the effect size distribution varies across causal variants).
Some phenotypes, such as schizophrenia or height, are highly heritable and highly polygenic \cite[]{sullivan2003schizophrenia,yang2010common,oconnor2019extreme,trubetskoy2022mapping}.
Causal effects are larger when a trait is more heritable, and smaller when it is more polygenic.
As for the distribution of causal effects relative to their allele frequencies, it is often investigated through a single parameter, usually called $\alpha$ or $S$, to model the effect of negative selection on complex traits whereby variants with lower frequencies are expected to have higher causal effect sizes \cite[]{speed2012improved}.
In this model, the expected phenotypic variance explained by a genetic variant is proportional to $\big[f(1-f)\big]^\alpha$, where $f$ is the allele frequency of this variant.
Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$. 
These methods (non-exhaustively) include GCTA\cite{yang2011gcta} ($h^2$), BOLT-REML\cite{loh2015contrasting} ($h^2$ and $p$), LD Score regression\cite{bulik2015ld} ($h^2$), FINEMAP\cite{benner2016finemap} (per-variant $p$, also called posterior inclusion probabilities PIPs, used for fine-mapping), HESS\cite{shi2016contrasting} (local $h^2$), LDAK-SumHer\cite{speed2019sumher} ($h^2$ and $\alpha$), S-LD4M\cite{oconnor2019extreme} ($p$), GRM-MAF-LD\cite{schoech2019quantification} ($\alpha$), SuSiE\cite{wang2020simple} (PIPs), SBayesS\cite{zeng2021widespread} ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR\cite{johnson2021estimation} (local $p$).

@@

Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a small polygenicity (Figure S24).
Note that these $h^2$ estimates (and later the $r^2$ estimates) have not been transformed to the liability scale (i.e.\ are on the observed scale).
Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure S25).
Most phenotypes have an estimated $\alpha$ between -1.0 and -0.4 with a mode at -0.65 (Figure S26).
As for the inferred predictive performance $r^2$ (from the Gibbs sampler of LDpred2-auto, Methods), they are highly consistent with the predictive performance in the test set; only for standing height are they overestimated (Figures \ref{fig:ukbb_r2} and S27). Heritability estimates for height are probably overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature (also see Section ``Application to height'').

@@

LDpred2-auto was originally developed for building polygenic scores \cite[]{prive2020ldpred2}.
Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP heritability (both genome-wide and more locally), polygenicity (and per-variant probabilities of being causal, also known as posterior inclusion probabilities, PIPs, useful for fine-mapping), and the selection-related parameter $\alpha$.
We remind readers that LDpred2 can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large,ding2022polygenic}.
We also present a new way to infer the out-of-sample predictive performance $r^2$ of the resulting PGS, assuming the target sample has the same genetic ancestry as the GWAS used for training.
Others have proposed to estimate $r^2$ from a reference genotype set \cite[]{mak2017polygenic} or from an additional set of external GWAS summary statistics and LD reference \cite[]{pattee2020penalized,witteveen2022publicly}; here we only use the summary statistics that we input to LDpred2-auto.
Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf (assuming an infinitesimal architecture, i.e.\ $p=1$) generally providing lower predictive performance than LDpred2-grid (testing different values for parameter $p$) or LDpred2-auto (estimating $p$)\cite[]{prive2020ldpred2}.
We also obtain widespread signatures of negative selection with most $\alpha$ estimates between -1.0 and -0.4, consistent with previous findings \cite[]{zeng2021widespread}.
However, when looking at the heritability enrichments of several functional annotations for height, we obtain much smaller magnitudes with LDpred2-auto than with stratified LD Score regression (S-LDSC)\cite{finucane2018heritability}. For example, \citet{yengo2022saturated} report fold enrichments of more than 10x for e.g.\ coding and conserved variants, while we get less than 2x. 
This is partly due to LDpred2-auto estimates being more conservative as they are shrunk towards no enrichment (the prior), however we do use a very large sample size here so that the prior should not matter much.
Another possible reason comes from using variants that capture the causal effects by LD, while these tagging variants may be annotated differently from the causal variants, which could cause functional enrichments to be diluted \cite[]{zheng2022leveraging}.
We also note that this heritability partitioning is performed after running LDpred2-auto for each annotation independently, therefore, unlike S-LDSC, the LDpred2-auto heritability partitioning does not depend on the set of annotations used.

@@

Our proposed method has limitations. First, when power is low (i.e.\ when $N h^2 / p$ is low), estimates of $\alpha$ and $p$ become less reliable.
Therefore, we recommend using the 3-parameter model (with $\alpha$, when using \texttt{use\_MLE = TRUE}), but only when power is sufficient and when robustness is not an issue.
We also recommend using option \texttt{allow\_jump\_sign = FALSE} for robustness \cite[]{prive2021identifying}, and for getting accurate or conservative $p$ estimates. When still obtaining a large $p$ estimate ($>1\%$) with this option, we recommend rerunning LDpred2-auto without this option to get a less conservative estimate (cf.\ simulations).
However, LDpred2-auto estimates of $h^2$ and $r^2$ seem always reliable, except for height for which they are probably overestimated.
We think this is likely due to assortative mating, which causes GWAS effects to be inflated \cite[]{border2022assortative,herzig2023model}, which causes our estimates to be inflated as well. 
Second, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics.
Future work could focus on correcting these issues.
Third, when using external GWAS summary statistics, it is often beneficial to regularize the LD matrix (via parameter \texttt{coef\_shrink}, especially when using the extended model and sampling), however it leads to an increased estimation for e.g.\ $r^2$ and $h^2$.
Future work could focus on correcting these estimates, and also on identifying the best shrinkage regularization coefficient to apply based on e.g.\ some distance metric (mismatch) between GWAS summary statistics and the LD reference used.
Fourth, because we use a limited set of variants as input for LDpred2, causal variants identified by LDpred2-auto are probably tagging variants that are highly correlated with unobserved causal variants close by. 
LDpred2-auto can also miss some causal variants when they are poorly tagged by the set of variants used.
Future work will focus on scaling LDpred2-auto to using more variants to alleviate this limitation; currently, the main issue is the large size of the LD matrix that grows quadratically with the number of variants used.
For now, to identify causal variants, one can concentrate on small regions of the genome, where LDpred2-auto can be rerun using a much more dense set of variants, as typically done in fine-mapping analyses (cf.\ Results section ``Local heritability and polygenicity'').

@@

\noindent A tutorial on running LDpred2 with R package bigsnpr (for both prediction and inference), \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}

>>>>

Most traits and diseases in humans are heritable.
What differs is the genetic architecture of each trait that can be parameterized by three key terms: the heritability (i.e.\ the proportion of phenotypic variation explained by genetics), the polygenicity (i.e.\ the fraction of genomic variants that have a non-zero effect on the trait), and the causal effect distribution (i.e.\ how the effect size distribution varies across causal variants).
Some phenotypes, such as schizophrenia or height, are highly heritable and highly polygenic \cite[]{sullivan2003schizophrenia,yang2010common,oconnor2019extreme,trubetskoy2022mapping}.
Causal effects are larger when a trait is more heritable, and smaller when it is more polygenic.
As for the distribution of causal effects relative to their allele frequencies, it is often investigated through a single parameter, usually called $\alpha$ or $S$, to model the effect of negative selection on complex traits whereby variants with lower frequencies are expected to have higher causal effect sizes \cite[]{speed2012improved}.
In this model, the expected phenotypic variance explained by a genetic variant is proportional to $\big[f(1-f)\big]^\alpha$, where $f$ is the allele frequency of this variant.
Many methods have been developed to estimate the SNP heritability (referred to as $h^2$ for brevity) and polygenicity ($p$), either globally for the whole genome or locally for specific regions of the genome, as well as $\alpha$. 
These methods (non-exhaustively) include GCTA\cite{yang2011gcta} ($h^2$), BOLT-REML\cite{loh2015contrasting} ($h^2$ and $p$), LD Score regression\cite{bulik2015ld} ($h^2$), FINEMAP\cite{benner2016finemap} (per-variant $p$, also called posterior inclusion probabilities PIPs, used for fine-mapping), HESS\cite{shi2016contrasting} (local $h^2$), LDAK-SumHer\cite{speed2019sumher,speed2020evaluating} ($h^2$ and $\alpha$), S-LD4M\cite{oconnor2019extreme} ($p$), GRM-MAF-LD\cite{schoech2019quantification} ($\alpha$), SuSiE\cite{wang2020simple} (PIPs), SBayesS\cite{zeng2021widespread} ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR\cite{johnson2021estimation} (local $p$).
Not all methods have the same modeling assumptions; for example, LDAK-SumHer assumes a different prior than SBayesS and LDpred2-auto, and does not estimate all the same parameters. It can estimate the SNP heritability and $\alpha$. However, it cannot estimate the polygenicity nor the per-variant probabilities of being causal (since it inherently assumes an infinitesimal model, i.e.\ $p = 1$). Moreover, since it does not sample effects, it cannot be used to estimate the predictive performance $r^2$ with the new formula we propose in this paper.

@@

Consistent with simulations, inferred SNP heritability $h^2$ estimates from LDpred2-auto closely match with those from LD Score regression, while generally being more precise, especially for phenotypes with a small polygenicity (Figure S24).
Note that these $h^2$ estimates (and later the $r^2$ estimates) have not been transformed to the liability scale (i.e.\ are on the observed scale).
Most phenotypes have an estimated polygenicity $p$ between 0.001 and 0.04; these have therefore a very polygenic architecture, but not an infinitesimal one (Figure S25).
Most phenotypes have an estimated $\alpha$ between -1.0 and -0.3 with a mode at -0.65 (Figure S26).
As for the inferred predictive performance $r^2$ (from the Gibbs sampler of LDpred2-auto, Methods), they are highly consistent with the predictive performance in the test set; only for standing height are they overestimated (Figures \ref{fig:ukbb_r2} and S27). Heritability estimates for height are probably overestimated as well since we use similar formulas for estimating $h^2$ and $r^2$ (Methods), and because the SNP heritability estimate $h^2$ for standing height is higher than values reported in the literature (also see Section ``Application to height'').

@@

LDpred2-auto was originally developed for building polygenic scores \cite[]{prive2020ldpred2}.
Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the SNP heritability (both genome-wide and more locally), polygenicity (and per-variant probabilities of being causal, also known as posterior inclusion probabilities, PIPs, useful for fine-mapping), and the selection-related parameter $\alpha$.
We remind readers that LDpred2 can also be used to infer the uncertainty of individual polygenic scores \cite[]{ding2022large,ding2022polygenic}.
We also present a new way to infer the out-of-sample predictive performance $r^2$ of the resulting PGS, assuming the target sample has the same genetic ancestry as the GWAS used for training.
Others have proposed to estimate $r^2$ from a reference genotype set \cite[]{mak2017polygenic} or from an additional set of external GWAS summary statistics and LD reference \cite[]{pattee2020penalized,witteveen2022publicly}; here we only use the summary statistics that we input to LDpred2-auto.
Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf (assuming an infinitesimal architecture, i.e.\ $p=1$) generally providing lower predictive performance than LDpred2-grid (testing different values for parameter $p$) or LDpred2-auto (estimating $p$)\cite[]{prive2020ldpred2}.
We also obtain widespread signatures of negative selection with most $\alpha$ estimates between -1.0 and -0.3 with a mode at -0.65, consistent with previous findings from SBayesR where they find a median S (same as  $\alpha$) of -0.578 (SD: 0.096) \cite[]{zeng2021widespread}.
In that paper, they found consistent results with BayesR, although the original BayesR publication reported weaker effects with a median of -0.37 (SD: 0.11) \cite[]{zeng2018signatures}. In \citet{schoech2019quantification}, they found an average $\alpha$ of -0.38 (SE: 0.02) with GRM-MAF-LD, but they also noted that LDAK estimates were upward biased by 0.4 (originally centered at -0.25 \cite[]{speed2020evaluating}, suggesting a likely value would be close to -0.65, which is what we obtain here).

However, when looking at the heritability enrichments of several functional annotations for height, we obtain much smaller magnitudes with LDpred2-auto than with stratified LD Score regression (S-LDSC)\cite{finucane2018heritability}. For example, \citet{yengo2022saturated} report fold enrichments of more than 10x for e.g.\ coding and conserved variants, while we get less than 2x. 
This is partly due to LDpred2-auto estimates being more conservative as they are shrunk towards no enrichment (the prior), however we do use a very large sample size here so that the prior should not matter much.
Another possible reason comes from using variants that capture the causal effects by LD, while these tagging variants may be annotated differently from the causal variants, which could cause functional enrichments to be diluted \cite[]{zheng2022leveraging}.
We also note that this heritability partitioning is performed after running LDpred2-auto for each annotation independently, therefore, unlike S-LDSC, the LDpred2-auto heritability partitioning does not depend on the set of annotations used.

@@

Our proposed method has limitations. First, when power is low (i.e.\ when $N h^2 / p$ is low), estimates of $\alpha$ and $p$ become less reliable.
Therefore, we recommend using the 3-parameter model (with $\alpha$, when using \texttt{use\_MLE = TRUE}), but only when power is sufficient and when robustness is not an issue.
We also recommend using option \texttt{allow\_jump\_sign = FALSE} for robustness \cite[]{prive2021identifying}, and for getting accurate or conservative $p$ estimates. When still obtaining a large $p$ estimate ($>1\%$) with this option, we recommend rerunning LDpred2-auto without this option to get a less conservative estimate (cf.\ simulations).
However, LDpred2-auto estimates of $h^2$ and $r^2$ seem always reliable, except for height for which they are probably overestimated.
We think this is likely due to assortative mating, which causes GWAS effects to be inflated \cite[]{border2022assortative,herzig2023model}, which causes our estimates to be inflated as well. 
Second, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics.
Future work could focus on correcting these issues.
Third, when using external GWAS summary statistics, it is often beneficial to regularize the LD matrix (via parameter \texttt{coef\_shrink}, especially when using the extended model and sampling), however it leads to an increased estimation for e.g.\ $r^2$ and $h^2$.
Future work could focus on correcting these estimates, and also on identifying the best shrinkage regularization coefficient to apply based on e.g.\ some distance metric (mismatch) between GWAS summary statistics and the LD reference used.
Fourth, because we use a limited set of variants as input for LDpred2, causal variants identified by LDpred2-auto are probably tagging variants that are highly correlated with unobserved causal variants close by. 
LDpred2-auto can also miss some causal variants when they are poorly tagged by the set of variants used.
Future work will focus on scaling LDpred2-auto to using more variants to alleviate this limitation.
LDpred2-auto runtime currently depends on the number of causal variants (or, equivalently, the polygenicity). When the polygenicity is close to 0, it runs linearly with the number of variants. When the polygenicity is close to 1, it runs quadratically. However, the polygenicity is almost always lower than 0.1 for human traits and diseases (cf.\ UKBB results here), which makes running LDpred2-auto efficient (we report runtimes always under one hour, even with HapMap3+ variants).
Currently, the main issue is the large size of the LD matrix that grows quadratically with the number of variants used.
For now, to identify causal variants, one can concentrate on small regions of the genome, where LDpred2-auto can be rerun using a much more dense set of variants, as typically done in fine-mapping analyses (cf.\ Results section ``Local heritability and polygenicity'').

@@

\noindent The official tutorial on running LDpred2 with R package bigsnpr (for both prediction and inference), \url{https://privefl.github.io/bigsnpr/articles/LDpred2.html}
